The Soulgem Paper

### 2011-09-24 soulgem.tex
Second draft of Theoretical Framework section.

Sociologists and organizational scientists have studied the concept of roles in social groups over decades. We take their findings as the starting points for the discussion below, acknowledging them where appropriate.

@@

A \textbf{\emph{role}} is a set of expectations of behaviour placed on a person given their position, status, history, or preference \cite{Turner1956,Gordon1976,Rizzo1970}. This entails that roles are \emph{social constructions}: in a real sense, a ``developer'' is whatever the relevant social group determines that a developer should be. There is no archetypal developer. Upon learning that a colleague is a developer, all the members of the organization place a number of expectations on said colleague. Perhaps they expect the developer to spend as much time as possible writing production code, to be responsive to user requests for assistance, to validate the code produced by other developers, or to self-manage. Some colleagues may have no expectations other than that the developer should strive for the well-being of the organization. Other colleagues, especially if they interact with the developer on a regular basis, may have many more expectations, and these may be far more concrete. People outside the formal organization, but with connections or interactions to it, may hold other expectations on her behaviour. And the developer herself, the target of all these expectations, likely has a number of additional expectations on her own conduct: the activities she feels she should be doing to play her part in the organization.

@@

A \textbf{\emph{position}} or a \textbf{\emph{title}}, in turn, is an official statement of a person's goals and responsibilities. A role is not a position \cite{Turner1956}, though they often carry the same labels: one can play an architectural role and hold the position of Software Architect. A person's title plays a part in determining what she will feel compelled to do, and what she believes she is expected to do, but it is only one element of her role. Roles are enacted, positions are occupied (several people could fulfill the same role, despite their different positions; a person can fulfill several roles while holding a single position). For instance, one could carry an architectural role without holding an Software Architect position, and one could hold a Software Architect position without performing any of the activities that others would expect from an architect.

We can see the positions in an organization, and their accompanying lists of responsibilities, as the explicit manifestations of what the fraction of the members of the organization that have the power to set those positions and responsibilities expect from their colleagues. In that sense they are a part of the total set of expectations placed upon organizational roles; a powerful part (especially if a position's responsibilities are tied to incentives and job evaluations), but still only a part.

As indicated above, we often use \textbf{\emph{labels}} to refer to roles. A fitting label helps us convey the essence of the role to our interlocutors. But as we will see some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms (in many organizations there are people who are expected to ``keep the place together,'' for example). We must then remember that neither the label nor the formal position make the role.

\subsection{Roles and culture}

Of course, the larger culture plays an important part in setting and steering our expectations. In some domains the culture is sufficiently strong to establish an interpretation of a role throughout the population. As we mention in the Introduction, in the medical domain, a combination of intensive training, codes of ethics, professional societies, and familiarization in person and through popular culture leads us to expect certain behaviours and activities from a doctor, and it leads the doctor to become attuned to the expectations placed upon her. We find it shocking if a doctor deviates from these expectations significantly. In our own domain, a similar combination of forces (education, past experiences, popular culture) allows us to make some general assumptions about software development roles that we can reasonably expect will hold in most cases. For example, we would be surprised if a tester did not spend at least part of his time preparing or running tests. And so, thanks to this larger culture, a newcomer to the test department will likely assume that he is expected to prepare and run tests, and his peers and supervisors will indeed have these expectations placed upon him, and this is what he will spend his time doing, thereby reinforcing the general idea of what testers do.

However, in our domain, the forces from the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist. As the evidence of our case study shows below, significant disagreements exist even within the same software organizations. Determining the reasons for this disagreement is beyond the scope of this paper; we speculate that the relative immaturity of the software industry, the wide variety of fields in which it functions, and the low barriers of entry to the profession all contribute to the phenomenon. But whatever the reason, the lack of conceptual uniformity in software development roles has implications for both research and practice, which we discuss in the coming pages.

\subsection{Dysfunction and role evolution}

In theory, we can use the definition of roles as sets of expectations of behaviour to build a conceptual map of roles in an organization: we find what every member of the organization expects from everyone else, and the aggregate of these expectations determines the roles that all members enact. But the members of the organization do not always act according to the expectations placed on them. The disagreement between expectation and actual behaviour is a major source of dysfunction in the organization.

Several reasons explain these disagreements between expectation and performance. The expectation can be unrealistic (``testers should find all the bugs in our code''), the performance may fall short of reasonable demands due to poor skill or carelessness, the recipient of the expectation may be unaware of it (a failure to coordinate or communicate), or the expectation could be in conflict with other expectations placed on the same role or on the same person.

Poor performance in the satisfaction of expectations is commonplace, but so is a performance that goes beyond the original expectation. If these disagreements recur over time, we undergo a shift in our expectations: we learn to expect the performance that is common to our experience, our ``new normal.'' And therefore, since roles are conformed by sets of expectations, and these expectations change over time, role definitions fluctuate in any social group based on past performance.

@@

\item Therefore, roles are \emph{social constructs}. There is no official definition of what a certain role is.

\item A set of expectations (a role) does not need to be satisfiable. When it is not, the individual enacting the role experiences \emph{role strain}.

@@

\item Expectations are partly set by the larger \emph{culture}. However, in our domain, the culture has not enforced a singular view of the roles in software organizations.

\item Disagreements between expectation and performance are a source of organizational \emph{dysfunction}.

\item Performance that diverges from expectations, either positively or negatively, affects future expectations. Therefore, past performance in an organization causes \emph{evolutions} in its role definitions.
\end{itemize}

Although most of these points are well established in the sociological literature, to our knowledge they had not been previously examined under the particular characteristics of software organizations. We provide empirical evidence to support them in the following pages.

% TODO: Add something about specialization with growth in this section?

>>>>

The following framework is distilled from the work of sociologists and organizational scientists. To our knowledge this is the first attempt at analyzing these concepts in the context of software development work.

@@

A \textbf{\emph{role}} is a set of expectations of behaviour placed on a person given their position, status, history, or preference \cite{Turner1956,Gordon1976,Rizzo1970}. This entails that roles are \emph{social constructions} \cite{Berger1967}: in a real sense, a ``developer'' is whatever the relevant social group determines that a developer should be. There is no prototypical developer. Upon learning that a colleague is a developer, all the members of the organization place a number of expectations on said colleague. Perhaps they expect the developer to spend as much time as possible writing production code, to be responsive to user requests for assistance, to validate the code produced by other developers, or to self-manage. Some colleagues may have no expectations other than that the developer should strive for the well-being of the organization. Other colleagues, especially if they interact with the developer on a regular basis, may have many more expectations, and these may be far more concrete: to adhere to the specified API or to put some test hooks in place. People outside the formal organization, but with connections or interactions to it, may hold other expectations on her behaviour. And the developer herself, the target of all these expectations, likely has a number of additional expectations on her own conduct: the activities she feels she should be doing to play her part in the organization.

@@

A \textbf{\emph{position}} or a \textbf{\emph{title}}, in turn, is an official statement of a person's goals and responsibilities. A role is not a position \cite{Turner1956}, though they often carry the same labels: one can play an architectural role and hold the position of Software Architect. A person's title plays a part in determining what she will feel compelled to do, and what she believes she is expected to do, but it is only one element of her role. For instance, one could enact an architectural role without holding an Software Architect position, and one could hold a Software Architect position without performing any of the activities that others would expect from an architect.

Positions and their responsibilities are often the explicit articulation of what a powerful fraction of the organization (the managers) expect from their colleagues. In that sense they are a part of the total set of expectations placed upon organizational roles; an important part (especially if tied to incentives and job evaluations), but still only a part.

As indicated above, we often use \textbf{\emph{labels}} to refer to roles. A fitting label helps us convey the essence of the role to our interlocutors. But some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms. We must remember that neither the label nor the formal position completely define the role.

\subsection{Institutionalized roles}

Culture plays an important part in setting and steering our expectations. In some domains the culture is sufficiently strong to establish an interpretation of a role throughout the population: the role has been institutionalized \cite{DiMaggio1991}. In some domains, a combination of intensive training, codes of ethics, professional societies, and familiarization lead us to expect certain behaviours and activities from a role. In our domain, these institutional forces allow us to make some general assumptions about software development roles that we can reasonably expect will hold in most cases. For example, we might be surprised if a tester did not spend at least part of his time preparing or running tests. And so, thanks to this process of institutionalization, a newcomer to the test department will likely assume that he is expected to prepare and run tests, his peers and supervisors will indeed place these expectations upon him, and this is what he will spend his time doing, thereby reinforcing the general idea of what testers do.

However, the forces from the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist. As the evidence of our case study shows, significant disagreements exist even within the same software organizations. Determining the reasons for this disagreement is beyond the scope of this paper; we speculate that the relative immaturity of the software industry, the wide variety of fields in which it functions, and the low barriers of entry to the profession all contribute to the phenomenon. Whatever the reason, we make no judgment regarding the convenience of this lack of uniformity in role definitions, but we believe it should be explored in greater detail.

\subsection{Role evolution}

The members of an organization may begin setting their expectations based on institutional roles, but they do not always act according to the expectations placed on them. Several reasons explain this disagreement. In the case of performance that fail to meet expectations, it could be that the expectation is unrealistic (``testers should find all the bugs in our code''), that the performer lacks skill or care, that the performer is unaware of the expectation (a failure to coordinate or communicate), or that the expectation is in conflict with other expectations (role strain).

It is easy to find examples of poor performance, but so is it to find performance that goes beyond the original expectation. In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal.'' They adapt their expectations, and since roles are conformed by these expectations, role definitions evolve in any social group based on past performance, despite their institutional status.

Another reason for role evolution arises from changes in the expectations that are needed to meet the organization's goals. An involvement in a new market niche may bring the need to coordinate the old products with the new one, to transfer expertise, and so on. One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971}. With growth comes specialization, and specialization consists of narrower role definitions. Therefore, other things being equal, we expect larger organizations to have a greater degree of role specialization.

@@

\item Therefore, roles are \emph{social constructs}. There is no prototypical definition of what a certain role is.

@@

\item Expectations are partly set by our \emph{institutions}, that is, by the rules, norms, and culture prevalent in our society.

\item Performance that diverges from expectations, either positively or negatively, affects future expectations. Therefore, past performance in an organization causes a role \emph{evolution}.

\item Organizational growth brings new requirements into the organization, which also leads to role evolution. In particular, growth tends to cause role \emph{specialization}.
\end{itemize}
